### 2022-07-25 src/text/1-introduction.tex
Extend introduction

%% Gamma-ray astronomy
Very high energy cosmic gamma-ray reaching Earth interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based gamma-ray astronomy relies on this phenomenon to detect the
primary gamma-ray photons and measure their direction and energy.
It covers the energy range from fews tens of GeV up to the PeV.
Two main instrument categories exist \citep{2015CRPhy..16..610D}.
Imaging Atmospheric Cherenkov Telescopes (IACT) make images of atmospheric showers
by detecting the Cherenkov radiation emitted by the cascading charged particles and
use these images to reconstruct the properties of the incident particle.

Particle samplers detect directly particles from the tail of the shower when it reaches
the ground. These instruments have very large field-of-view, large duty-cycle but higher
energy threshold and usually have lower signal to noise ratios compared to IACTs.

%% Context
Ground based gamma-ray astronomy has been historically structured
by experiments run by independent collaborations relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been very successful so far, it does not
permit easy combination of data from several instruments and is therefore
a limitation for interoperability of existing facilities.

@@

%% Context : formats
In practice, the data-reduction workflow of IACTs is remarkably similar.
After data calibration, shower events are reconstructed and
gamma/hadron separation is applied to build lists of gamma-ray like events.
The latter are then used to derive scientific results, such as spectra, sky maps
or light curves, taking into account the Instrument Response Functions (IRF).
The information in this high data level is independent on
the data reduction, and eventually of the detection technique. This implies,
for example, that data from IACT and particle samplers can be represented
within the same model. The efforts to prototype a format usable by various instruments
converged in the so-called \textit{Data Format for Gamma-ray Astronomy}
initiative~\citep{gadf_proc, gadf_universe}, abbreviated in
\texttt{gamma-astro-data-format} (GADF). The latter proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of gamma-like events with their measured quantites
(energy, direction, arrival time) and a parametrisation of the response of the
system. (see Sec.~\ref{ssec:gammapy-data} and Sec.~\ref{ssec:gammapy-irf} for
more information).

>>>>

%% Gamma-ray astronomy
Gamma-ray astronomy is a rather young field of research. The \gammaray~
range of the electromagnetic spectrum provides us with insight to the
most energetic process in the universe such as surroundings of
black holes and remnants of supernova explosions. As in other
branches of astronomy \gammarays can be observed by both
satellite as well as ground based instruments.
Very high energy cosmic gamma-rays reaching Earth interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based gamma-ray astronomy relies on this phenomenon to detect the
primary gamma-ray photons and measure their direction and energy.
It covers the energy range from fews tens of GeV up to the PeV.
There are two main categories of instrument \citep{2015CRPhy..16..610D}.
Imaging Atmospheric Cherenkov Telescopes (IACT) make images of atmospheric showers
by detecting the Cherenkov radiation emitted by the cascading charged particles and
use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution.

Water Cherenkov observatories detect directly particles from the tail of the
shower when it reaches the ground. These instruments have very
large field-of-view, large duty-cycle but higher energy threshold and
usually have lower signal to noise ratios compared to IACTs.

%% Context
Ground based gamma-ray astronomy has been historically structured
by experiments run by independent collaborations relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been very successful so far, it does not
permit easy combination of data from several instruments and is therefore
a limitation for interoperability of existing facilities. Especially
because the different detection techniques have complementary
properties.

@@

%% Context : formats
In practice, the data-reduction workflow of all gamma-ray observatories
is remarkably similar.
After data calibration, shower events are reconstructed and
gamma/hadron separation is applied to build lists of gamma-ray like events.
The latter are then used to derive scientific results, such as spectra, sky maps
or light curves, taking into account the specific instrument response functions (IRF).
The information in this high data level is independent on
the data reduction, and eventually of the detection technique. This implies,
for example, that data from IACT and particle samplers can be represented
within the same model. The efforts to prototype a format usable by various instruments
converged in the so-called \textit{Data Format for Gamma-ray Astronomy}
initiative~\citep{gadf_proc, gadf_universe}, abbreviated in
\texttt{gamma-astro-data-format} (GADF). The latter proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of gamma-like events with their measured quantites
(energy, direction, arrival time) and a parametrisation of the response of the
system. (see Sec.~\ref{ssec:gammapy-data} and Sec.~\ref{ssec:gammapy-irf} for
more information).
